Coherent optical association of a single molecule

### 2020-09-22 paper.tex
Update on Overleaf.

Such ultracold molecular association takes advantage of the much developed cooling and trapping techniques for atoms  as a starting point. To overcome the challenges of small wavefunction overlap and the large release of binding energy of converting atoms to deeply-bound molecules, a  two-step approach has been established to first associate atom pairs into weakly-bound molecules, and then transfer the molecules from this single internal state to a desired rovibrational and electronic state~\cite{Danzl2008, Ni2008,Lang2008, Takekoshi2014, Molony2014, Park2015, Guo2016, Kondov2019, Voges2020}.
So far, all of such association processes  utilized a magnetic Feshbach scattering resonance and have been applied to bialkali molecules. The only exception is Sr$_2$ where narrow linewidth excited states are available and optical association can be driven~\cite{Reinaudi2012,Stellmer2012}. The requirement of a Feshbach resonance to enhance atom-to-molecule wavefunction overlap or the narrow excited state lines limits the generality of the association technique. 
%to more diverse molecular species.

Here, we demonstrate coherent association of an atom pair to a weakly bound molecule using a two-photon optical Raman transfer via an electronic excited state, without the use of a Feshbach resonance nor a narrow excited state. Our scheme is based on a theoretically determined vibrational state of the electronic excited state $c^3\Sigma^+$ that has the best transfer Rabi coupling to photon scattering ratio. To further increase this ratio and reduce technical requirements such as intensity stability, we choose an initial and final state where the matrix elements with the excited state are as balanced as possible. This approach minimizes the reliance on system specific properties and can therefore be more generally applied to creating other molecular species or larger molecules atom-by-atom with full quantum state control. %The choice of the states is general and can be applied to associate other molecular species or large molecules atom-by-atom with full quantum state control. 
(also this result represents full quantum-state control of single [neutral] molecules, the second time ever (first being our NaCs Feshbach molecules)) %This approach minimizes the reliance on system specific properties and can therefore be more generally applied to creating other molecules.
%electronic the best initial hyperfine combination of the atoms to use and best single photon frequency that balances coupling/transfer efficiency and photon scattering. Using these parameters, we demonstrate coherent association of a electronic ground-state molecule from two atoms via an optical Raman transition in an optical tweezer.

@@

\begin{figure*}
  \includegraphics[height=5cm]{fig1.pdf}
  \caption{Optical creation of single molecule from single atoms in tweezer.
    (A) Schematics of the optical transition from an atom pair to a weakly bound molecule.
    The initial state is the relative motional ground state between the two atoms
    and the final state is the first molecular bound state.
    The transition is driven by a pair of laser frequencies matching the binding energy
    of the molecule.
    The lasers are detuned from an excited molecular state in the $c^3\Sigma$ potential
    by $\Delta$ in order to suppress the scattering during the transfer.
    (B) Comparison between using a weakly bound and a deeply bound excited state
    as intermediate state for the Raman transition.
    The deeply bound excited state (upper half $v=0$) has a smaller Raman Rabi frequency ($\Omega_{Raman}$)
    compared to the weakly bound excited state (lower half $v=63$) at a given detuning.
    However, the scattering rate ($\Gamma_{scatter}$) is also much lower,
    which results in a larger Raman Rabi frequency to scattering rate ratio.
    (C) Enhancement of short range wavefunction.
    The large scattering length for the $Na(2,2),Cs(3,3)$ state creates an interaction shift
    comparable to the axial trapping frequency.
    This causes a significant change in the relative wavefunction especially at short
    intranuclear distance ($R$).
    Compared to other spin states with weaker interaction,
    the wavefunction at short distance ($R<100\mathrm{\AA}$) is significantly enhanced.
    \label{f-theory}
  }
\end{figure*}

@@

%While optical Raman transfers of atoms to molecules utilizing an electronic excited state have been shown previously, the transfers were incoherent  where the resulting molecules were lost~\cite{Wynar2000,Rom2004}.

The challenge of optical Raman transfers of atoms to molecules lies with achieving a high enough Raman Rabi frequency  due to the wavefunction size mismatch with the excited molecular state. While pioneering experiments used weakly bound molecular excited states in the Raman transition to increase the Raman Rabi frequency, this choice of state also increases the scattering during the transfer process to render it incoherent, resulting in molecule loss~\cite{Wynar2000,Rom2004}. 
%Previous experiments used weakly bound molecular excited states in the Raman transition to increase the Raman Rabi frequency. However, this choice of state also increases the scattering during the transfer process for two reasons.
In general, scattering can be reduce by detuning away from resonance. However, we are constrained by the multi-level of molecules. 
Firstly, since for a typical molecular potential the spacing between states
decreases with increasing vibrational levels,
a weakly bound excited molecular state limits the maximum usable detuning and therefore the effectiveness of scattering reduction by increasing the detuning for the Raman transition. Secondly, due to size similarity(?), the weakly bound molecular ground state has a strong coupling with the atomic excited state which contributes significantly to the total scattering rate. The rate of such a scattering process is larger for the weakly bound states, and is proportional to $1/\delta_{thresh}^2$,
where $\delta_{thresh}$ is the detuning from the dissociation threshold. (Should we also mention scattering of the atomic initial state too? This should affect the Raman process as well)
% This further increases the scattering rate for a weakly bond excited state.
Based on these considerations, we calculated both the Raman Rabi frequency and the scattering rate at different detunings (full result is shown in SM). As seen in figure (1....), we found a distinct regime from what have been explored previously, namely, despite a smaller Raman Rabi frequency at deeply bound intermediate states, the Rabi frequency to scattering ratio is much higher as compared to a weakly bound intermediate state. As a result, we use a deeply bound molecular excited state ($v'=0$ of $c^3\Sigma^+$) to drive the Raman transition. (mention the trend, also v=12 will work as similarly as v=0)

@@

In addition to the intermediate state, the choice of the initial and final states are critical for both fundamental and technical reasons. A key difference in molecular association via a Raman process compared to Raman transfer between hyperfine states in atoms is that the matrix elements between the ground states and the excited state are greatly unbalanced. In particular, the matrix element between the molecule state and the excited state, $ \Omega_m $ is much larger than the matrix element between the free atom state and the excited state, $ \Omega_a $. To understand the effect of this imbalance, we consider a 3 level model, and ignore the presence of multiple excited states. The Raman Rabi Rate, $ \Omega_a\Omega_m / 2\Delta$ depends on both matrix elements, but the scattering rate predominantly comes from the molecular state, $ \Gamma_e \Omega_m^2 / 2\Delta^2$, where $ \Gamma_e $ is the excited state linewidth and $ \Delta $ is the single photon detuning. We choose the power of the Raman beams to be equal (this gives the smallest scattering), there is an additional factor of 2 in the scattering rate, since the molecular state scatters off of both Raman beams at roughly the same single photon detuning. Thus, the ratio between the Raman Rabi frequency and the scattering rate, $ \Omega_a/\Omega_m \times \Delta/\Gamma_e $, depends on the ratio of the two matrix elements and how far detuned the laser is from the transition in units of the linewidth. Thus, a small $ \Omega_m/\Omega_a $ ratio is desired to be able to drive this transition coherently at smaller detunings where the Raman Rabi frequency will be larger. In addition to this fundamental reason, there is also a technical advantage for having a $ \Omega_m/\Omega_a $ ratio to be as small( KW: I think small, but open question whether we should always talk about $\Omega_m/\Omega_a$ or $ \Omega_a/\Omega_m$, KN: maybe $ \Omega_a/\Omega_m$ is better because it sounds more natural that we want to make it large? but either way.) as possible. The position of the resonance depends on the laser power predominantly through the AC Stark shift on the molecular state, $ \Omega_m^2 / 2\Delta $. The ratio of the AC Stark shift to the Raman Rabi frequency is $ \Omega_m / \Omega_a $, Thus, the laser needs to be stabilized to better than the inverse of this ratio, $ \Omega_a/\Omega_m $, to fluctuate by less than a linewidth. This becomes technically easier when $ \Omega_m/\Omega_a $ is smaller.
%(for our choice of states, this mean intensity fluctuation of less than 1\% or 0.1\%. KW: I mentioned this in next paragraph, does it still need to be mentioned here?)

This ratio can be made smaller by increasing the coupling of the atomic state with the excited molecular state. Due to the small size of the molecular wavefunction, the coupling between the ground atomic state and the excited molecular state is approximately proportional to the value of the relative atomic wavefunction at short distance within the molecular potential. In additional to the confinement, this value is related to the interaction between the two atoms. For states with a large scattering length (positive or negative), the phase shift in the relative wavefunction between the atoms can significantly increase the short range wavefunction (add figure?). The increase in the coupling is proportional to (quote/cite Olive's equation?). For our system, among the stable spin combinations, 4422 and 3311 both have small scattering lengths of $\abs{a_{4422}}= 30.4a_0, \abs{a_{3311}} = 13.7a_0 $ respectively and 3322 has a large and negative scattering length of $ \abs{a_{3322}} = 693.8a_0 $ (interaction shift $\approx$ binding?)~\cite{Hood2019}. In addition to the increased atomic coupling, $ \Omega_a $, with the 3322 hyperfine combination, coupled channel calculations show that the most weakly bound molecular state that is predominantly this spin combination also has reduced coupling, $ \Omega_m $, with the excited state when compared with the 4422 and 3311 bound states. Therefore, using an initial 3322 hyperfine combination results in a $ \Omega_m/\Omega_a$ ratio of about 20 instead of a ratio of about 300 with the 4422 or 3311 combinations. This more favorable ratio also relaxes the technical requirement of intensity stability of $0.3 \% $ for the 4422 combination compared to $ 5\% $ for the 3322 combination.  Thus, we use the 3322 spin combination as our initial state and drive to the first bound state for the 3322 spin combination. (We might want to mention that those ratios are for 80 kHz spherical trap if we want to give more details about the coupled channel calculation anywhere...)

>>>>

Such ultracold molecular association takes advantage of the much developed cooling and trapping techniques for atoms  as a starting point. To overcome the challenges of small wavefunction overlap and the large release of binding energy of converting atoms to deeply-bound molecules, a  two-step approach has been established to first associate atom pairs into weakly-bound molecules, and then transfer the molecules from this single internal state to a desired rovibrational and electronic state~\cite{Danzl2008, Ni2008,Lang2008, Takekoshi2014, Molony2014, Park2015, Guo2016, Kondov2019, Voges2020}.
So far, all of such association processes  utilized a magnetic Feshbach scattering resonance and have been applied to bialkali molecules. The only exception is Sr$_2$ where narrow linewidth excited states are available and optical association can be driven~\cite{Reinaudi2012,Stellmer2012}. The requirement of a Feshbach resonance to enhance atom-to-molecule wavefunction overlap or the narrow excited state lines limits the generality of the association technique. % to more diverse species.

Here, we demonstrate coherent association of an atom pair to a weakly bound molecule using a two-photon optical Raman transfer via an electronic excited state, without the use of a Feshbach resonance nor a narrow excited state. The resulting single molecule is in a single internal quantum state and predominately in its motional ground state.  Our scheme is based on a theoretically determined vibrational state of the electronic excited state $c^3\Sigma^+$ that has the best transfer Rabi coupling to photon scattering ratio. To further increase this ratio and reduce technical requirements such as intensity stability, we choose an initial and final state where the matrix elements with the excited state are as balanced as possible. This approach minimizes the reliance on system specific properties and can therefore be more generally applied to creating other molecular species or larger molecules atom-by-atom with full quantum state control. 
%(also this result represents full quantum-state control of single [neutral] molecules, the second time ever (first being our NaCs Feshbach molecules))

%This approach minimizes the reliance on system specific properties and can therefore be more generally applied to creating other molecules.
%electronic the best initial hyperfine combination of the atoms to use and best single photon frequency that balances coupling/transfer efficiency and photon scattering. Using these parameters, we demonstrate coherent association of a electronic ground-state molecule from two atoms via an optical Raman transition in an optical tweezer.

@@

\begin{figure*}
  \includegraphics[height=5cm]{fig1.pdf}
  \caption{Optical creation of single molecule from single atoms in tweezer.
    (A) (should label $v'=0$ and $v'=63$) Schematics of the optical transition from an atom pair to a weakly bound molecule.
    The initial state is the relative motional ground state between the two atoms
    and the final state is the first molecular bound state.
    The transition is driven by a pair of laser frequencies matching the binding energy
    of the molecule.
    The lasers are detuned from an excited molecular state in the $c^3\Sigma$ potential
    by $\Delta$ in order to suppress the scattering during the transfer.
    (B) Comparison between using a weakly bound and a deeply bound excited state
    as intermediate state for the Raman transition.
    The deeply bound excited state (upper half $v=0$) has a smaller Raman Rabi frequency ($\Omega_{Raman}$)
    compared to the weakly bound excited state (lower half $v=63$) at a given detuning.
    However, the scattering rate ($\Gamma_{scatter}$) is also much lower,
    which results in a larger Raman Rabi frequency to scattering rate ratio.
    (C) Enhancement of short range wavefunction.
    The large scattering length for the $Na(2,2),Cs(3,3)$ state creates an interaction shift
    comparable to the axial trapping frequency.
    This causes a significant change in the relative wavefunction especially at short
    intranuclear distance ($R$).
    Compared to other spin states with weaker interaction,
    the wavefunction at short distance ($R<100\mathrm{\AA}$) is significantly enhanced.
    \label{f-theory}
  }
\end{figure*}

@@

%While optical Raman transfers of atoms to molecules utilizing an electronic excited state have been shown previously, the transfers were incoherent  where the resulting molecules were lost~\cite{Wynar2000,Rom2004}. 
%For a given excited intermediate state,
%This can be understood by using a three level model. 
%However, we are constrained by the multi-level of molecules. 
%A key difference in molecular association via a Raman process compared to Raman transfer between hyperfine states in atoms is that the matrix elements between the ground states and the excited state are greatly unbalanced. In particular, the matrix element between the molecule state and the excited state, $ \Omega_m $ is much larger than the matrix element between the free atom state and the excited state, $ \Omega_a $. To understand the effect of this imbalance, we consider a 3 level model, and ignore the presence of multiple excited states.

The challenge of optical Raman transfers of atoms to molecules lies with achieving a high enough Raman Rabi frequency  due to the wavefunction size mismatch with the excited molecular state. While pioneering experiments used weakly bound molecular excited states in the Raman transition to increase the Raman Rabi frequency, such choice of state also increases the scattering during the transfer process to render it incoherent, resulting in molecule loss~\cite{Wynar2000,Rom2004}. 
Therefore, finding the largest Raman Rabi frequency to scattering ratio that enables population transfer with a minimal lost is the first key task.

In a simplified model of a three-level system (as shown in Fig.~\ref{f-theory}), the initial atomic state, an excited intermediate state, and the target weakly-bound molecular state in the ground electronic potential are coupled by two photons, $\Omega_a$ and $\Omega_m$.  %(Ultimately, we calculate with contribution of all vibrational states of the excited electronic potential, $c^3\Sigma^+$)
The Raman Rabi Rate, $ \Omega_a\Omega_m / 2\Delta$ depends on both the matrix element between the atomic state and the excited state, $ \Omega_a  $ and the matrix element between the molecular state and the excited state, $ \Omega_m $. Unlike in atoms, the two matrix elements, $ \Omega_m$ and $ \Omega_a$ are greatly imbalanced. Since $ \Omega_m $ is generally much larger than $ \Omega_a $ due to the wavefunction mismatch between the atomic state and the excited molecular state, the scattering rate predominantly comes from the molecular state, $ \Gamma_e \Omega_m^2 / 2\Delta^2$, where $ \Gamma_e $ is the excited state linewidth and $ \Delta $ is the single photon detuning. We choose the power of the Raman beams to be equal (which gives the largest Raman Rabi rate), so there is an additional factor of 2 in the scattering rate, since the molecular state scatters off of both Raman beams at roughly the same single photon detuning. Thus, the ratio between the Raman Rabi frequency and the scattering rate, $ \Omega_a/\Omega_m \times \Delta/\Gamma_e $, depends on the ratio of the two matrix elements and how far detuned the laser is from the transition in units of the linewidth. To ensure a coherent process, larger detuning is preferred, but if it's too large, the Raman Rabi rate may become too small. With the multiple excited vibrational states present in a molecular potential, the total scattering rates becomes a sum of the scattering rates over all the states. Since for a typical molecular potential the spacing between states
decreases with increasing vibrational levels,
a weakly bound excited molecular state limits the maximum usable detuning and therefore the effectiveness of the scattering reduction by increasing the detuning for the Raman transition. Furthermore, due to similarity in wavefunction size, the weakly bound molecular ground state has a strong coupling with the atomic excited state which contributes significantly to the total scattering rate. The rate of such a scattering process is larger for weakly bound molecular excited states, and is proportional to $1/\delta_{thresh}^2$,
where $\delta_{thresh}$ is the detuning from the dissociation threshold. (Should we also mention scattering of the atomic initial state too? This should affect the Raman process as well)
% This further increases the scattering rate for a weakly bond excited state.
Based on these considerations, % a weakly bound excited molecular state may not be the best choice for the intermediate state in the Raman transfer. Thus, 
we calculated both the Raman Rabi frequency and the scattering rate at different detunings (full result is shown in SM). As seen in figure (1....), we found a distinct regime from what have been explored previously, namely, despite a smaller Raman Rabi frequency at deeply bound intermediate states, the Rabi frequency to scattering ratio is much higher as compared to a weakly bound intermediate state. As a result, we use a deeply bound molecular excited state ($v'=0$ of $c^3\Sigma^+$) to drive the Raman transition. (mention the trend, also v=12 will work as similarly as v=0)

@@

%In addition to the intermediate state, the choice of the initial and final states are critical for both fundamental and technical reasons.  Thus, a small $ \Omega_m/\Omega_a $ ratio is desired to be able to drive this transition coherently at smaller detunings where the Raman Rabi frequency will be larger. In addition to this fundamental reason, there is also a technical advantage for having a $ \Omega_m/\Omega_a $ ratio to be as small( KW: I think small, but open question whether we should always talk about $\Omega_m/\Omega_a$ or $ \Omega_a/\Omega_m$, KN: maybe $ \Omega_a/\Omega_m$ is better because it sounds more natural that we want to make it large? but either way.) as possible. The position of the resonance depends on the laser power predominantly through the AC Stark shift on the molecular state, $ \Omega_m^2 / 2\Delta $. The ratio of the AC Stark shift to the Raman Rabi frequency is $ \Omega_m / \Omega_a $, Thus, the laser needs to be stabilized to better than the inverse of this ratio, $ \Omega_a/\Omega_m $, to fluctuate by less than a linewidth. This becomes technically easier when $ \Omega_m/\Omega_a $ is smaller.
%(for our choice of states, this mean intensity fluctuation of less than 1\% or 0.1\%. KW: I mentioned this in next paragraph, does it still need to be mentioned here?)

In addition to the intermediate state, the choice of the initial and final states are critical for both fundamental and technical reasons. As previously discussed, the ratio of the Raman Rabi rate to the scattering rate is proportional to the $ \Omega_a/\Omega_m $ ratio. The smaller this ratio is, the farther the laser needs to be detuned to achieve the same Raman Rabi rate to scattering rate ratio. This ratio can be made smaller by increasing the coupling of the atomic state with the excited molecular state. Due to the small size of the molecular wavefunction, the coupling between the ground atomic state and the excited molecular state is approximately proportional to the value of the relative atomic wavefunction at short distance within the molecular potential. In additional to the confinement, this value is related to the interaction between the two atoms. For states with a large scattering length (positive or negative), the phase shift in the relative wavefunction between the atoms can significantly increase the short range wavefunction (add figure?). The increase in the coupling is proportional to (quote/cite Olive's equation?). For our system, among the stable spin combinations, 4422 and 3311 both have small scattering lengths of $\abs{a_{4422}}= 30.4a_0, \abs{a_{3311}} = 13.7a_0 $ respectively and 3322 has a large and negative scattering length of $ \abs{a_{3322}} = 693.8a_0 $ (interaction shift $\approx$ binding?)~\cite{Hood2019}. In addition to the increased atomic coupling, $ \Omega_a $, with the 3322 hyperfine combination, coupled channel calculations show that the most weakly bound molecular state that is predominantly this spin combination also has reduced coupling, $ \Omega_m $, with the excited state when compared with the 4422 and 3311 bound states. Therefore, using an initial 3322 hyperfine combination results in a $ \Omega_a/\Omega_m$ ratio of about 0.05 instead of a ratio of about 0.003 with the 4422 or 3311 combinations. In addition to reducing how far the laser needs to be detuned from the transition, a larger $ \Omega_a/\Omega_m$ ratio also relaxes the intensity stability requirement. The position of the resonance depends on the laser power predominantly through the AC Stark shift on the molecular state, $ \Omega_m^2 / 2\Delta $. The ratio of the AC Stark shift to the Raman Rabi frequency is $ \Omega_m / \Omega_a $, Thus, the laser needs to be stabilized to better than the inverse of this ratio, $ \Omega_a/\Omega_m $, to fluctuate by less than a linewidth. Thus, instead of a required intensity stability of $0.3 \% $ for the 4422 combination, the 3322 combination only requires the intensity to be stabilized to $ 5\% $.  Thus, we use the 3322 spin combination as our initial state and drive to the first bound state for the 3322 spin combination. (We might want to mention that those ratios are for 80 kHz spherical trap if we want to give more details about the coupled channel calculation anywhere...)
